A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex
First draft of RIPPLES numbers summary

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
% Slightly risky, this, as we could get asked to incorporate 
% novel recombs found by RIPPLEs? This could be done if we could get 
% Viridian sample IDs for their list.
been thoroughly characterised. 
The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.

>>>>

[Rough first draft]
This estimate for effective recombination in SARS-CoV-2 is much lower 
than previous studies have suggested.
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
% accessed SARS-CoV-2 genomes

% Although multiple studies have 
% published lists of putatitive novel recombinant 
% sequences~\cite{,Turakhia2022,Li2024CovRecomb} 
% these are for GISAID data and not trival to reconcile with 
% the Viridian data or the specific recombination events in the ARG.

% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
% recombination events, 
% RecombinHunt just looks at X lineages (I think?)

To assess the evidence for these putative novel recombinants

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
been thoroughly characterised.

Although multiple studies have 
published lists of putatitive novel recombinant 
sequences~\cite{Varabyou2021rapid,Turakhia2022,Li2024CovRecomb} 
these are for GISAID data and not trival to reconcile with 
the Viridian data or the specific recombination events in the ARG.

% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
% accessed SARS-CoV-2 genomes
% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
% recombination events, 
% RecombinHunt just looks at X lineages (I think?)

To assess the evidence for these putative novel recombinants

The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.
